Using structural and evolutionary information to correct RNA pyrosequencing errors in non-coding RNAs

### 2013-05-24 JCB/methods_JCB.tex
Stated problem more explicitly

\subsubsection{Energy Contribution.}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, using values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. neither $\Gb\Ub$, $\Ub\Gb$, $\Cb\Gb$, $\Gb\Cb$, $\Ab\Ub$ nor $\Ub\Ab$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed. For instance, multiloops do not consider base-specific contributions, and therefore their consideration would constitute a criterion for preferring a sequence over another. Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution.}
The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the isostericity contribution to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, $0$ corresponding to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution exponentially favors sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.
\TODOYann{Avoid duplicate reference to Stombaugh et al}

@@

Let $s$ be an RNA sequence, $S$ a reference structure, and $m\geq 0$ a desired number of mutations. 
We are interested in  the probability that a given position contains a specific nucleotide, over all sequences having at most $M$ mutations from $s$
%  under the SCFG derivation $S$ 
(formally $\mathbb{P}(s_i = x\mid s,\Omega, \Struct,M)$). 
We define a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990} to compute this
 probability, based on two sets of dynamic programming equations. 
%from the two  functions $\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$.

The former, defined in Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec_A}-\eqref{eq:Z_rec_C}, is analogous to an \emph{inside} computation. Considering an interval of the sequence/secondary structure, it consists in the accumulated contributions of any possible sequences that have suitable Hamming distance within the interval. It is therefore similar to a partition function, i.e. the sum of Boltzmann factors over all sequences within $[i,j]$, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$), within 
$m$ mutations of $s$. The latter, defined by Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec_A}-\eqref{eq:Y_rec_D},
 computes the \emph{outside} algorithm,   
the partition function over sequences within $m$ mutations of $s$ outside the interval (restricted to two intervals $[0,i]\cup[j,n-1]$), 
knowing  that flanking inner positions $(i+1,j-1)$ contain nucleotides $a$ and $b$ respectively. A suitable combination of these terms, computed as shown in Equations~\eqref{eq:combine_A}-\eqref{eq:combine_C}, gives the total weight of every possible sequences that support a given base-pair (or unpaired position) and, in turn, the probability of seeing a specific base at a given position.

@@

\subsection{Complexity Considerations}
Clearly, Equations~\eqref{eq:Z_rec_A}-\eqref{eq:Z_rec_C} and~\eqref{eq:Y_rec_A}-\eqref{eq:Y_rec_D} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and structure lengths, storing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ may be accessed in constant time after a simple precomputation of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations due to the explicit distribution of the number of mutations.

In principle, $\Theta(m\cdot n^2)$ terms, identified by $(m,i,j)$ triplets, should be computed.
However, a close inspection of the recurrences reveals that the computation can be safely restricted to a subset of intervals $(i,j)$.
For instance, the inside algorithm only requires computing intervals $[i,j]$ that do not break any base-pair, and whose next position $j+1$ is either past the end of the sequence, or is base-paired prior to $i$. Similar constraints hold for the outside computation, resulting in a drastic limitation of the combinatorics of required computations, dropping from $\Theta(n^2)$ to $\Theta(n)$ the number of terms that need to be computed and stored. Consequently the overall complexity of the algorithm is $\Theta(n\cdot(|\Omega|+m^2))$ arithmetic operations and $\Theta(n\cdot(|\Omega|+m))$ memory.

>>>>

\subsubsection{Energy Contribution.}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, using values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. neither $\Gb\Ub$, $\Ub\Gb$, $\Cb\Gb$, $\Gb\Cb$, $\Ab\Ub$ nor $\Ub\Ab$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed. For instance, multiloops do not consider base-specific contributions, and therefore their consideration would constitute a criterion for preferring a sequence over another. Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity Contribution.}
The concept of isostericity score is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the isostericity contribution to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, $0$ corresponding to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution exponentially favors sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.
\TODOYann{Avoid duplicate reference to Stombaugh et al}

@@

Let $s$ be an RNA sequence, $S$ a reference structure, and $m\geq 0$ a desired number of mutations. 
We are interested in  the probability that a given position contains a specific nucleotide, over all sequences having at most $M$ mutations from $s$. Formally, let $\mathcal{D}_{s,M}=\{s'\;|\;\delta(s,s')\le M\}$ be the set of admissible sequences, one aims at computing the probabilities
\begin{equation}
\mathbb{P}(s_i = x\mid s,\Omega, \Struct,M) = \frac{\sum_{\substack{s'\in\mathcal{D}_{s,M}\\\text{s.t. }s'_{i}=x}}B(s')}{\sum_{\substack{s''\in\mathcal{D}_{s,M}}}B(s'')}
\end{equation}

Clearly, the number of sequences in $\mathcal{D}_{s,M}$ grows exponentially with the sequence length, therefore one cannot realistically rely on an exhaustive enumeration to compute the mutational profile. To work around this issue, we propose a linear-time variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990} to compute this probability, based on two sets of dynamic programming equations. 
%from the two  functions $\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$. 
The former, defined in Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec_A}-\eqref{eq:Z_rec_C}, is analogous to an \emph{inside} computation. Considering an interval of the sequence/secondary structure, it consists in the accumulated contributions of any possible sequences that have suitable Hamming distance within the interval. It is therefore similar to a partition function, i.e. the sum of Boltzmann factors over all sequences within $[i,j]$, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$), within 
$m$ mutations of $s$. The latter, defined by Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec_A}-\eqref{eq:Y_rec_D},
 computes the \emph{outside} algorithm,   
the partition function over sequences within $m$ mutations of $s$ outside the interval (restricted to two intervals $[0,i]\cup[j,n-1]$), 
knowing  that flanking inner positions $(i+1,j-1)$ contain nucleotides $a$ and $b$ respectively. A suitable combination of these terms, computed as shown in Equations~\eqref{eq:combine_A}-\eqref{eq:combine_C}, gives the total weight of every possible sequences that support a given base-pair (or unpaired position) and, in turn, the probability of seeing a specific base at a given position.

@@

\subsection{Complexity Considerations}
Using dynamic programming, Equations~\eqref{eq:Z_rec_A}-\eqref{eq:Z_rec_C} and~\eqref{eq:Y_rec_A}-\eqref{eq:Y_rec_D} can be computed in linear time and memory. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and structure lengths, storing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ may be accessed in constant time after a simple precomputation of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations due to the explicit distribution of the number of mutations.

In principle, $\Theta(m\cdot n^2)$ terms, identified by $(m,i,j)$ triplets, should be computed.
However, a close inspection of the recurrences reveals that the computation can be safely restricted to a subset of intervals $(i,j)$.
For instance, the inside algorithm only requires computing intervals $[i,j]$ that do not break any base-pair, and whose next position $j+1$ is either past the end of the sequence, or is base-paired prior to $i$. Similar constraints hold for the outside computation,  following from the linearity of the outside recurrence (i.e. the computation of the outside term for a given excluded structure only relies on the computation of another, strictly larger, structure).

These properties drastically limit the combinatorics of required computations, dropping from $\Theta(n^2)$ to $\Theta(n)$ the number of terms that need to be computed and stored. Consequently the overall complexity of the algorithm is $\Theta(n\cdot(|\Omega|+M^2))$ arithmetic operations and $\Theta(n\cdot(|\Omega|+M))$ memory.
